---
title: "computeTEW"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
   %\VignetteIndexEntry{computeTEW}
   %\VignetteEncoding{UTF-8}
   %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
library(StormR)
library(terra)
```

The `computeTEW()` function allows computing topographic exposure (TEW)
to wind for each cell of a regular grid (i.e., a raster) for a given
tropical cyclone or set of tropical cyclones. The `product="Profiles"`
and `product="PixProfiles"` argument allows producing exposure fields
during the lifespan of the cyclone at a temporal resolution of 1 hour.
The `computeTEW()` function also allows to compute three associated
summary statistics: the maximum topographic exposure for each cell
during the cyclone (`product="Max"`), the power dissipation index
(`product="Mean"`) and the topographic exposure to wind computed for
each pixel with the direction at maximal sustained speed along the
cyclone (`product="Summary"`). Then the `plotTEW()` and the
`writeRast()` functions can be used to visualise and export the output.
In the following example we use the `test_dataset` and `test_datadtm`
provided with the package to illustrate how cyclone track data can be
used to compute, plot, and export topographic exposure profiles or
summary statistics.

### Computing and plotting computeTEW products

#### Exctracting storm

We can compute the behaviour of winds generated by the topical cyclone
Pam (2015) near Vanuatu. First, track data are extracted as follows:

```{r}
sds <- defStormsDataset()
st <- defStormsList(sds = sds, loi = "Vanuatu", names = "PAM", verbose = 0, maxDist = 50)
plotStorms(st)
```

#### Exctracting topography

Then a Digital Terrain Model (DTM) is essential to compute topographic
exposure to wind (TEW). In this tutorial, DTM is from Copernicus DEM30 data and is
extracted thanks to the `openEO` package as follows :

```{r connection, eval=FALSE}
library(openeo)
# connect to the backend and authenticate (an account is requirred)
connection = connect(host = "https://openeofed.dataspace.copernicus.eu")
login()

# get the process collection to use the predefined processes of the back-end
p = processes()

#define wanted data
data = p$load_collection(id = 'COPERNICUS_30',
                         spatial_extent = list(west=168.0798, 
                                               south=-17.8480,
                                               east=168.6373,
                                               north=-17.3925),
                         bands = c("DEM"))


result = p$save_result(data = data, format = "GTiff")
```

```{r save tif file, eval = FALSE}
# save the tif file
compute_result(graph = result, output_file = "../filepath/name.tif")
```

```{r}
dtm <- rast("../inst/extdata/test_datadtm.tif")
```

#### Topographic exposure profiles

Using track data and the `computeTEW()` function with the
`product="Profiles"` or `product="PixProfiles"` argument we can generate
topographic exposure fields at any time as follows:

```{r}
pf <- computeTEW(st, dtm, product = "Profiles", verbose = 0)
pf
```

The product `Profiles` compute topographic exposure to wind (TEW) from one and
only direction taken at maximal speed for the entire location of
interest whereas `PixProfiles` compute topographic exposure to wind from
one direction per pixel. The first one consider an homogeneous wind
direction but the second one compute a more accurate picture.

The function returns a `SpatRaster` one raster for each observation or
interpolated observation. Rasters' names follow the following
terminology, the name of the storm in capital letters, "Exposure", and
the index of the observation, separated by underscores. Note that actual
observations have entire indices (e.g., 41, 42, ...) while interpolated
observation have decimals (e.g., 41.1, 41.2, ...). Most tropical cyclone
track data sets are based on observations gathered every 3 or 6 hours.
Therefore, to be able to compute wind fields every 1 hour, the
observations are interpolated.

Topographic exposure profiles at the 41^th^ observation can be plotted
as follows:

```{r}
plotTEW(st,dtm, pf$PAM_TEW_41)

```

#### Summary Statistics

The `computeTEW()` function can compute the different `products`
(i.e., `"Profiles"`, `"PixProfiles"`, `"Max"`, `"Mean"`, `"Summary"`)
either separately or together. Here, we compute all three summary
statistics together as follows:

```{r}
ss <- computeTEW(st, product = c("Max", "Mean", "Summary"), verbose = 0)
ss
```

The `computeTEW()` function returns a `SpatRaster` object with
three rasters: one for the maximum topographic exposure (`"Max"`), one
for the mean topographic exposure (`"Mean"`), and one for the
topographic exposure computed at maximum speed for each pixel
(`"Summary"`).

```{r}
names(ss)
```

The maximum topographic exposure can be plotted as follows:

```{r}
plotTEW(st, ss$PAM_TEW_Max)
```

The mean topographic exposure can be plotted as follows:

```{r}
plotTEW(st, ss$PAM_TEW_Mean)
```

The summary of topographic exposure can be plotted as follows:

```{r}
plotTEW(st, ss$PAM_TEW_Summary)
```


#### Dynamic plot

`plotTEW` function also provide (See ExtractStorms vignette) a dynamic plot.
Here is an example based on the same parameters as above.

```{r, eval = FALSE}
plotTEW(st, ss$PAM_TEW_Summary, dynamicPlot = TRUE)
```
**warning : ne fonctionne peut être pas pour le moment**

### Exporting computeTEW products

The `computeTEW()` function returns rasters stored in a `SpatRaster`
object than can be exported either in ".tiff" or ".nc" (NetCDF) formats using
the `writeRast()` function. Here, we export the summary of topographic exposure in
the working directory as follows:

```{r}
writeRast(ss$PAM_TEW_Summary)
```
