---
title: "computeTEW"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
   %\VignetteIndexEntry{computeTEW}
   %\VignetteEncoding{UTF-8}
   %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
bibliography: stormr.bib
---

```{r setup, include = FALSE}
devtools::load_all()
#library(StormR)
library(terra)
```

The `computeTEW()` function allows computing topographic exposure to wind (TEW)
@boose_hurricane_1994 for each cell of a regular grid (i.e., a raster) for a given
tropical cyclone or set of tropical cyclones. The `product="Profiles"`
and `product="PixProfiles"` argument allows producing exposure fields
during the lifespan of the cyclone.
The `computeTEW()` function also allows to compute three associated
summary statistics: the maximum and mean TEW for each cell
during the cyclone (`product="Max"` and `product="Mean"`, respectively) 
and the TEW computed for
each pixel with the direction at maximal sustained speed along the
cyclone (`product="Summary"`). 

Then the `plotTEW()` and the `writeRast()` functions can be used to visualize and export the output.

In the following example we use the `test_dataset` and `test_datadtm`
provided with the package to illustrate how cyclone track data can be
used to compute, plot, and export topographic exposure profiles or
summary statistics.

### Computing and plotting computeTEW products

#### Exctracting storm

We can compute the behaviour of winds generated by the topical cyclone
Pam (2015) near Vanuatu. First, track data are extracted as follows:

```{r}
sds <- defStormsDataset()
st <- defStormsList(sds = sds, loi = "Vanuatu", names = "PAM", verbose = 0, maxDist = 50)
plotStorms(st)
```

#### Exctracting topography

Then a Digital Terrain Model (DTM) is essential to compute topographic
exposure to wind (TEW). In this tutorial, DTM used is [Copernicus DEM30 data](dataspace.copernicus.eu/explore-data/data-collections/copernicus-contributing-missions/collections-description/COP-DEM) and is
extracted thanks to the [`openEO` package](https://open-eo.github.io/openeo-r-client/index.html) as follows :

```{r connection, eval=FALSE}
library(openeo)
# connect to the backend and authenticate (an account is requirred)
connection = connect(host = "https://openeofed.dataspace.copernicus.eu")
login()

# get the process collection to use the predefined processes of the back-end
p = processes()

#define wanted data
data = p$load_collection(id = 'COPERNICUS_30',
                         spatial_extent = list(west=168.310547, 
                                               south=-17.651056,
                                               east=168.422470,
                                               north=-17.573830),
                         bands = c("DEM"))


result = p$save_result(data = data, format = "GTiff")
```

```{r save tif file, eval = FALSE}
# save the tif file
compute_result(graph = result, output_file = "../filepath/name.tif")
```

```{r}
dtm <- rast("../inst/extdata/test_datadtm.tif")
```

Other DTM products should work seemlessly with StormR.

#### Topographic exposure profiles

Using track data and the `computeTEW()` function with the
`product="Profiles"` or `product="PixProfiles"` argument we can generate
topographic exposure fields at any time as follows:

```{r}
pf <- computeTEW(st, dtm, product = "Profiles", verbose = 0)
pf
```

The product `Profiles` compute TEW from one and
only one direction taken at maximal wind speed location for the entire location of
interest. This option should be used for small locations of interest, where 
wind direction is considered homogeneous.

On the other hand, `PixProfiles` option computes TEW from
one direction per pixel. This gives a more accurate picture for large scale areas 
where wind direction variations are important.

The function returns a `SpatRaster` one raster for each observation or
interpolated observation. Rasters' names follow the following
terminology, the name of the storm in capital letters, "TEW", and
the index of the observation, separated by underscores. Note that actual
observations have entire indices (e.g., 41, 42, ...) while interpolated
observation have decimals (e.g., 41.1, 41.2, ...). Most tropical cyclone
track data sets are based on observations gathered every 3 or 6 hours.
Therefore, to be able to compute wind fields every 1 hour, the
observations are interpolated. This naming is coherent with `spatialBehaviour` and
`temporalBehaviour` functions.

Topographic exposure profiles at the 41^th^ observation can be plotted
as follows:

```{r}
plotTEW(st,dtm, pf$PAM_TEW_41)

```

#### Summary Statistics

The `computeTEW()` function can compute the different `products`
(i.e., `"Profiles"`, `"PixProfiles"`, `"Max"`, `"Mean"`, `"Summary"`)
either separately or together. "Summary" statistics are products integrated over 
the lifespan of the storm. Here, we compute all three summary
statistics together as follows:

```{r}
ss <- computeTEW(st, dtm,product = c("Max", "Mean", "Summary"), verbose = 0)
ss
```

The `computeTEW()` function returns a `SpatRaster` object with
three rasters: 

- one for the maximum TEW (`"Max"`): TEW is computed for each time step
using "Profiles", and the the maximum value is return for each pixel.
- one for the mean TEW (`"Mean"`, also taken as default value): TEW is 
computed for each time step using "Profiles", 
and the the mean value is return for each pixel.
- one for the maximum speed for each pixel TEW (`"Summary"`): TEW is computed 
using wind direction associated with maximum wind speed over the whole lifespan 
of the storm for each pixel.

```{r}
names(ss)
```

The maximum topographic exposure can be plotted as follows:

```{r}
plotTEW(st,dtm, ss$PAM_TEW_Max)
```

The mean topographic exposure can be plotted as follows:

```{r}
plotTEW(st, dtm,ss$PAM_TEW_Mean)
```

The summary of topographic exposure can be plotted as follows:

```{r}
plotTEW(st,dtm, ss$PAM_TEW_Summary)
```


#### Dynamic plot

`plotTEW` function also provide (See ExtractStorms vignette) a dynamic plot.
Here is an example based on the same parameters as above.

```{r, eval = FALSE}
plotTEW(st, dtm,ss$PAM_TEW_Summary, dynamicPlot = TRUE)
```
**warning : ne fonctionne peut être pas pour le moment**

### Exporting computeTEW products

The `computeTEW()` function returns rasters stored in a `SpatRaster`
object than can be exported either in ".tiff" or ".nc" (NetCDF) formats using
the `writeRast()` function. Here, we export the summary of topographic exposure in
the working directory as follows:

```{r}
writeRast(ss$PAM_TEW_Summary)
```

## References
