---
title: "computeTEW"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
   %\VignetteIndexEntry{computeTEW}
   %\VignetteEncoding{UTF-8}
   %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
bibliography: stormr.bib
---

```{r setup, include = FALSE}
library(StormR)
library(terra)
```

The `computeTEW()` function allows computing topographic exposure to
wind (TEW) @boose_hurricane_1994 for each cell of a regular grid (i.e.,
a raster) for a given tropical cyclone or set of tropical cyclones. The
`product="TEW1wd"` and `product="TEW"` argument allows producing
exposure fields during the lifespan of the cyclone. The `computeTEW()`
function also allows to compute three associated summary statistics: the
maximum and mean TEW for each cell during the cyclone
(`product="TEW1wdMax"` and `product="TEW1wdMean"`, respectively) and the
TEW computed for each pixel with the direction at maximal sustained
speed along the cyclone (`product="TEWIntegrated"`).

Then the `plotTEW()` and the `writeRast()` functions can be used to
visualize and export the output.

In the following example we use the `test_dataset` and `test_datadtm`
provided with the package to illustrate how cyclone track data can be
used to compute, plot, and export topographic exposure profiles or
summary statistics.

### Computing and plotting computeTEW products

#### Exctracting storm

We can compute the behaviour of winds generated by the topical cyclone
Pam (2015) near Vanuatu. First, track data are extracted as follows:

```{r}
sds <- defStormsDataset()
st <- defStormsList(sds = sds, loi = "Vanuatu", names = "PAM", verbose = 0, maxDist = 50)
plotStorms(st)
```

#### Exctracting topography

Then a Digital Terrain Model (DTM) is essential to compute topographic
exposure to wind (TEW). In this tutorial, DTM used is [Copernicus DEM30
data](dataspace.copernicus.eu/explore-data/data-collections/copernicus-contributing-missions/collections-description/COP-DEM)
and is extracted thanks to the [`openEO`
package](https://open-eo.github.io/openeo-r-client/index.html) as
follows :

```{r connection, eval=FALSE}
library(openeo)
# connect to the backend and authenticate (an account is requirred)
connection <- connect(host = "https://openeofed.dataspace.copernicus.eu")
login()

# get the process collection to use the predefined processes of the back-end
p <- processes()

# define wanted data
data <- p$load_collection(
  id = "COPERNICUS_30",
  spatial_extent = list(
    west = 168.982086,
    south = -19.002400,
    east = 169.348755,
    north = -18.614363
  ),
  bands = c("DEM")
)


result <- p$save_result(data = data, format = "GTiff")
```

```{r save tif file, eval = FALSE}
# save the tif file
compute_result(graph = result, output_file = "../filepath/name.tif")
```

Here, a sample DTM of Erromango Island included in `StormR` can be used:

```{r}
dtm <- rast(system.file("extdata","test_datadtm.tif",package = "StormR"))
```

Other DTM / elevation products should work seemlessly with StormR.

#### Extracting spatial profiles

Spatial profiles must be computed prior to run the `computeTEW()`
function.

```{r}
spProfiles <- spatialBehaviour(st, product = "Profiles", verbose = 0)
spProfiles
```

Other spatial profiles can be used as long as the layers of the
spatRaster respect the following terminology : the name of the storm in
capital letters, "Speed" or "Direction", and the indices of the
observation separated by underscores (e.g., "PAM_Speed_41",
"PAM_Direction_41", ...).

#### Topographic exposure to wind profiles

Using track data and the `computeTEW()` function with the
`product="TEW1wd"` or `product="TEW"` argument we can generate
topographic exposure fields at any time as follows:

```{r}
pf <- computeTEW(spProfiles, st, dtm, product = "TEW", verbose = 0)
pf
```

The product `TEW1wd` compute TEW from one and only one wind direction
taken at maximal wind speed location for the entire location of
interest. This option should be used for small locations of interest,
where wind direction is considered homogeneous.

On the other hand, `TEW` option computes TEW from one direction per
pixel. This gives a more accurate picture for large scale areas where
wind direction variations are important.

Multiple products can be computed at the same time, by simply providing
a vector of product names, e.g. `product = c("TEW1wd", "TEW")`.

The function returns a `SpatRaster` one raster for each observation or
interpolated observation. Rasters' names follow the following
terminology, the name of the storm in capital letters, "TEW", and the
index of the observation, separated by underscores. Note that actual
observations have entire indices (e.g., 41, 42, ...) while interpolated
observation have decimals (e.g., 41.1, 41.2, ...). Most tropical cyclone
track data sets are based on observations gathered every 3 or 6 hours.
Therefore, to be able to compute wind fields every 1 hour, the
observations are interpolated. This naming is coherent with
`spatialBehaviour` and `temporalBehaviour` functions.

Topographic exposure profiles at the 41^th^ observation can be plotted
as follows:

```{r}
plotTEW(st, pf$PAM_TEW_41, dtm = dtm)
```

#### Topographic exposure to wind at specified locations

Topographic exposure to wind can also be computed at any location of
interest.

This can be used in association with the `temporalBehaviour` function:

```{r}
df <- data.frame(x = c(169.097), y = c(-18.723))
rownames(df) <- c("Forest Plot")
TS <- temporalBehaviour(st, points = df, product = "TS", tempRes = 30, verbose = 0)
pfTemp <- computeTEW(TS, df, dtm, verbose = 0)
```

And you can then plot the time series of the topographic exposure to
wind at your location of interest as follows:

```{r}
plotTemporalTEW(pfTemp, "PAM")
```

#### Summary Statistics

In addition to TEW at all observations time, the `computeTEW()` function
can compute different "summary" `products`, either separately or
together. "Summary" statistics are products integrated over the lifespan
of the storm. Here, we compute all three integrated summary statistics
together as follows:

```{r}
ss <- computeTEW(spProfiles, st, dtm, product = c("TEW1wdMax", "TEW1wdMean", "TEWIntegrated"), verbose = 0)
ss
```

The `computeTEW()` function returns a `SpatRaster` object with three
rasters:

-   one for the maximum TEW (`"TEW1wd_Max"`): TEW is computed for each
    time step using "Profiles" and only one wind direction (associated
    with the maximum wind speed for each layer of "Profiles"), and the
    the maximum value is return for each pixel.
-   one for the mean TEW (`"TEW1d_Mean"`): TEW is computed for each time
    step using "Profiles", and only one wind direction (associated with
    the maximum wind speed for each layer of "Profiles"), and the the
    mean value is return for each pixel.
-   one for the maximum speed for each pixel TEW (`"TEW_Integrated"`):
    TEW is computed using wind direction associated with maximum wind
    speed over the whole lifespan of the storm for each pixel.

```{r}
names(ss)
```

The maximum topographic exposure can be plotted as follows:

```{r}
plotTEW(st, ss$PAM_TEW1wd_Max, dtm = dtm)
```

The mean topographic exposure can be plotted as follows:

```{r}
plotTEW(st, ss$PAM_TEW1wd_Mean, dtm = dtm)
```

The summary of topographic exposure can be plotted as follows:

```{r}
plotTEW(st, ss$PAM_TEW_Integrated, dtm = dtm)
```

#### Dynamic plot

`plotTEW` function also provide (See ExtractStorms vignette) a dynamic
plot. Here is an example based on the same parameters as above.

```{r}
plotTEW(st, ss$PAM_TEW_Integrated, dtm = dtm, dynamicPlot = TRUE)
```

### Exporting computeTEW products

The `computeTEW()` function returns rasters stored in a `SpatRaster`
object than can be exported either in ".tiff" or ".nc" (NetCDF) formats
using the `writeRast()` function. Here, we export the topographic
exposure to wind in the working directory as follows:

```{r}
writeRast(ss$PAM_TEW_Integrated)
```

### Example of application : computing the Exposure Vulnerability Index

The time-series layers produced by `computeTEW()` (e.g., one layer per
observation time from "`TEW`" or "`TEW1wd`" products) can be
post-processed to build advanced ecological proxies. For instance,
following the methodology defined by @mclaren_ev_2019, Exposure
Vulnerability (EV) can be computed to a specific storm or set of storms.

$EV = \frac{\sum_{i=1}^{n} (\text{MSW}_i \times \text{TEW}_i)}{n}$

It can be computed as follows :

```{r, include = FALSE}
dtm <- clamp(dtm, lower = 0.001, values = FALSE)
col <- colorRampPalette(c("darkgreen", "yellow", "red"))(100)
```

```{r}
# align msw on dtm
msw <- mask(project(spProfiles[[grep("_Speed", names(spProfiles))]], dtm), dtm)
# normalise
rg <- global(pf, c("min", "max"), na.rm = TRUE)
tew <- (pf - min(rg$min)) / (max(rg$max) - min(rg$min))
```

```{r}
ev <- app(msw * tew, mean, na.rm = TRUE)
```

High EV index values indicate critical vulnerability zones, highlighting
areas at peak risk for severe forest canopy damage and mechanical
disturbance.

The EV index can be plotted as follows :

```{r}
plot(ev, col = col)
```

This cumulative index can be an extended landscape-scale predictor for
long-term forest dynamics.

## References
