---
title: "DataSource"
output:
  pdf_document: default
  html_document: default
date: "`r Sys.Date()`"
---

```{r step up, include = FALSE}
#library(StormR)
```

Before running `StormR` functions users have to provide a tropical storm track
dataset as a ".nc" (NetCDF) file in which the location and some characteristics
of storms are given across their lifespan. This file is used to create a
`StormsDataset` object using the `defDatabase` function. By default, the
arguments of `defDatabase` function are set up to create a `StormsDataset`
object using the fields provided by USA agencies in the IBTrACS database
[International Best Track Archive for Climate
Stewardship](https://www.ncei.noaa.gov/products/international-best-track-archive)
(Knapp et al., 2010). This database provides a fairly comprehensive record of
worldwide tropical storms and cyclones with a 3-hours temporal resolution since
1841. Other databases can be used as long as the following fields are provided:

| **Field name** | **Description** | **Format/Units** | **Example** | **Type** |
|:--:|:----:|:-----------:|:-----:|:--:|
| `names` | Names of the storms | Capital letters | PAM | Mandatory |
| `seasons` | Years of observations | Calendar year | 2015 | Mandatory |
| `isoTime` | Date and time of observations | YYYY-MM-DD HH:mm:ss UTC | 2015-03-13 12:00:00 | Mandatory |
| `lon` | Longitude of the observations | Eastern Decimal degrees | 168.7 | Mandatory |
| `lat` | Latitude of the observations | Northern Decimal degrees | -17.6 | Mandatory |
| `msw` | Maximum sustained wind speed | $m.s^{-1}$ | 77 | Mandatory |
| `basin` | Name of the area where the storm originated. Traditionally divided into seven basins | IBTRaCS format \newline NA (North Atlantic) \newline EP (Eastern North Pacific) \newline WP (Western North Pacific) \newline NI (North Indian) \newline SI (South Indian) \newline SP (Southern Pacific) \newline SA (South Atlantic)| SP | Recommended$^{1}$|
| `rmw` | Radius of maximum winds: distance between the center of the storm and its band of strongest winds | $km$ | 20 | Recommended$^{2}$ |
| `pressure` | Central pressure | $pa$ | 91100 | Recommended$^{3}$ |
| `poci`| Pressure of the last closed isobar | $pa$ | 92200 | Recommended$^{3}$ |
| `sshs`| Saffir-Simpson hurricane wind scale rating based on `msw` (here in $m.s^{-1}$) | $-1 =$ Tropical depression (`msw` $< 18$) \newline $0 =$ Tropical storm ($18 \leq$ `msw` $< 33$) \newline $1 =$ Category 1 ($33  \leq$ `msw` $< 42$) \newline $2 =$ Category 2 ($42 \leq$ `msw` $< 49$) \newline $3 =$ Category 3 ($49 \leq$ `msw` $< 58$) \newline $4 =$ Category 4 ($58 \leq$ `msw` $< 70$) \newline $5 =$ Category 5 (`msw` $\ge 70$)| 5 | Optional$^{4}$  |

$^{1}$ This field let the user filter storms on a particular area. It can be
very usefull for large dataset if one is interested only in a specific basin. If
not provided, this filtering operation cannot be achieved. <br />

$^{2}$ Providing `rmw` allows better estimates of wind speed and direction. If
not provided `rmw` is approximated using an empirical formula derived from
Willoughby et al. (2006) : $R_m = 46.4e^{(-0.0155 \times v_m + 0.0169 \times
|\phi|)}$. <br />

$^{3}$ If these fields are not filled, both Holland and Boose models cannot be
performed (See Behaviour vignette)<br />

$^{4}$ The `sshs` field is optional, if not provided this field is automatically
filled using the `msw` field.

### Downloading a data source

If not available, a storm track data set has to be downloaded. In the following
example we illustrate how a storm track data set can be downloaded on the
IBTrACKS website. Different ".nc" (NetCDF) files containing storm track data
sets can be downloaded here:
<https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/netcdf/>.
Please, follow the IBTrACS citation recommendations when using these datasets
(including citing Knapp et al., 2010 and Knapp et al., 2018, and providing the
access date for the later). For example, track data for all the storms that
occurred over the last three years over the different basin world wide can be
download on the working directory as follows,

```{r, eval = F}
download.file(url = "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/netcdf/IBTrACS.last3years.v04r00.nc", destfile = "./IBTrACS_ALL_3yrs.nc")
```

### Creating a `StormsDataset`

Once a storm track data set is available a `StormsDataset` object can be created
using the `defDatabase` function as follows,

```{r, eval=F}
sds <- defDatabase(filename = "IBTrACS_ALL_3yrs.nc", verbose = FALSE)
str(sds)
length(sds@database$names)
```

This `StormsDataset` object contains track data for 344 storms. When using an IBTrACKS storm track data set, default arguments can be used. By default, field provided by USA agencies (starting with "usa_") are used, but other fields can be selected. For example, for the South Pacific basin ("SP"), we can choose to use data provided by the Australian bureau of meteorology (starting with "bom_") as follows,

```{r, eval = F}
sds <-defDatabase(filename = "IBTrACS_ALL_3yrs.nc",
                  fields = c(names = "name",
                             seasons = "season",
                             isoTime = "iso_time",
                             basin = "basin",
                             lon = "bom_lon",
                             lat = "bom_lat",
                             msw = "bom_wind",
                             rmw = "bom_rmw",
                             pressure = "bom_pres",
                             poci = "bom_poci"),
                  basin = "SP",
                  verbose = FALSE)
length(sds@database$names)
```

This `StormsDataset` object contains track data for 38 storms in the South
Pacific basin.

### Unit conversion

In the IBTrACS data sets speeds are given in knots ($knt$), distances in
nautical miles ($nm$), and pressure in millibar ($mb$) but the models
implemented in the `temporalBehaviour` and `spatialBehaviour` functions to
compute wind speed, wind direction, and summary statistics require data
following the international system of units (SI) or SI-derived units. Therefore,
by default, when using IBTrACS data sets, the `defDatabase` function converts
knots ($knt$) to meter per second ($m.s^{-1}$), nautical miles ($nm$) to
kilometres ($km$), and millibar ($mb$) to Pascal ($pa$). When using another
source of data than IBTrACS, the `unit_conversion` argument can be used to
convert the data to the desire units.The following lists the allowed conversions
available when using the `unit_conversion` argument

For `msw`: \newline
`knt_to_ms`: converts knot in meter per second (Default) \newline
`kmh_to_ms`: converts kilometer per hour in meter per second \newline
`mph_to_ms`: converts miles per hour in meter per second \newline
`None`: No conversion needed \newline


For `rmw`: \newline
`nm_to_ms`: converts nautical miles in kilometer (Default) \newline
`None`: No conversion needed \newline


For both `pressure` and `poci`: \newline
`b_to_pa`: converts bar in Pascal \newline
`mb_to_pa`: converts millibar in Pascal (Default) \newline
`atm_to_pa`: converts atmosphere in Pascal \newline
`psi_to_pa`: converts psi in Pascal \newline
`None`: No conversion needed \newline


These fields in `unit_conversion` argument are mandatory if they are also contained
in the fields of `fields` argument. If a field does not required any conversion,
the user must still specify `"None"`.

### Dataset example

This package provides a dataset `test_dataset` which gathers all the tropical
cyclones that occured nearby Vanuatu (resp. New Caledonia) from 2015 to 2016
(resp. 2020 to 2021 ). It represents the default `StormsDataset` used in
the `Storms` function if no other `StormsDataset` has been initialized
with the `defDatabase` function.


## Reference
 * Knapp, K. R., Kruk, M. C., Levinson, D. H., Diamond, H. J., & Neumann, C. J. (2010). The International Best Track Archive for Climate Stewardship (IBTrACS). Bulletin of the American Meteorological Society, 91(3), Article 3. https://doi.org/10.1175/2009bams2755.1

  * Knapp, K. R., H. J. Diamond, J. P. Kossin, M. C. Kruk, & Schreck, C. J. (2018) International Best Track Archive for Climate Stewardship (IBTrACS) Project, Version 4. [indicate subset used]. NOAA National Centers for Environmental Information. https://doi:10.25921/82ty-9e16 [2023-03-15]


#############################################



\textbf{- Lien vers page téléchargement IBTraCKS OKKK}
\textbf{- Lien vers page téléchargement autre BDD ???}
\textbf{- Code pour download OKKK}


In the following example, we will use the data from the IBTrACS that gather all
storms that occurred in the South Pacific Basin (IBTrACS.SP.v04r00.nc). It is
possible to download this file using:
```{r chunk, eval=FALSE}
download.file(url = "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/netcdf/IBTrACS.SP.v04r00.nc",destfile="IBTrACS_ALL_3yrs.nc")
```


## Define a StormsDataset
\textbf{- iniDataBase sur les 2 BDD téléchargées}
\textbf{- différence de gestion IBTRACKS car paramètres par défaut vs autre bdd ou il faut redéfinir les champs.}
\textbf{- dire qu'on a déjà une BDD par défaut de test déjà existant (e.g. ?IBTRACKS\_SP), pas besoin de faire un initDatabase OKKKK}
\textbf{- Gestion des unités ? OKKKKKK}

The `defDatabase` function let the user define a new dataset. This function
returns a `StormsDataset` object, especially designed to store all the
informations needed for the dataset (See Data sources section).
In what follows, we define a new `StormsDataset` based on the database we
downloaded right above and the default settings of `defDatabase` function.
```{r chunk2, eval = FALSE}

sds <- defDatabase(filename = "./test_ibtracs.nc",
                   fields = c(basin = "basin",
                              names = "name",
                              seasons = "season",
                              isoTime = "iso_time",
                              lon = "usa_lon",
                              lat = "usa_lat",
                              msw = "usa_wind",
                              sshs = "usa_sshs",
                              rmw = "usa_rmw",
                              pressure = "usa_pres",
                              poci = "usa_poci"),
                   basin = "SP",
                   seasons = c(1980, as.numeric(format(Sys.time(), "%Y"))),
                   unit_conversion = c(msw = "knt_to_ms",
                                       rmw = "nm_to_km",
                                       pressure = "mb_to_pa",
                                       poci = "mb_to_pa"),
                   verbose = TRUE)

```

Note:
If one is extracting data from IBTrACS databases, we strongly advise using the `usa_` fields as they are
the most filled and relevant.
The field `basin` here, will allow searching for storms that occurred only in the South Pacific Basin.
The field `seasons` here, will allow searching for storms that occurred between 1980 and the current year.



