---
title: "Data source"
output:
  html_document: default
  pdf_document: default
date: "`r Sys.Date()`"
---

```{r step up, include = FALSE}
#library(StormR)
```

Before running `StormR` functions users have to provide a tropical storm track
dataset as a ".nc" (NetCDF) file in which the location and some characteristics
of storms are given across their lifespan. This file is used to create a
`StormsDataset` object using the `defDatabase` function. By default, the
arguments of `defDatabase` function are set up to create a `StormsDataset`
object using the fields provided by USA agencies in the IBTrACS database
[International Best Track Archive for Climate
Stewardship](https://www.ncei.noaa.gov/products/international-best-track-archive)
(Knapp et al., 2010). This database provides a fairly comprehensive record of
worldwide tropical storms and cyclones with a 3-hours temporal resolution since
1841. Other databases can be used as long as the following fields are provided:

| **Field name** | **Description** | **Format/Units** | **Example** | **Type** |
|:--:|:----:|:-----------:|:-----:|:---:|
| `names` | Names of the storms | Capital letters | PAM | Mandatory |
| `seasons` | Years of observations | Calendar year | 2015 | Mandatory |
| `isoTime` | Date and time of observations | YYYY-MM-DD HH:mm:ss UTC | 2015-03-13 12:00:00 | Mandatory |
| `lon` | Longitude of the observations | Eastern decimal degrees | 168.7 | Mandatory |
| `lat` | Latitude of the observations | Northern decimal degrees | -17.6 | Mandatory |
| `msw` | Maximum sustained wind speed | $m.s^{-1}$ | 77 | Mandatory |
| `basin` | Name of the area where the storm originated. Traditionally divided into seven basins | NA (North Atlantic) <br /> EP (Eastern North Pacific) <br /> WP (Western North Pacific) <br /> NI (North Indian) <br /> SI (South Indian) <br /> SP (Southern Pacific) <br /> SA (South Atlantic)| SP | Recommended$^{1}$|
| `rmw` | Radius of maximum winds: distance between the centre of the storm and its band of strongest winds | $km$ | 20 | Recommended$^{2}$ |
| `pressure` | Central pressure | $pa$ | 91100 | Recommended$^{3}$ |
| `poci`| Pressure of the last closed isobar | $pa$ | 92200 | Recommended$^{3}$ |
| `sshs`| Saffir-Simpson hurricane wind scale rating based on `msw` (here in $m.s^{-1}$) | $-1 =$ Tropical depression (`msw` $< 18$) <br /> $0 =$ Tropical storm ($18 \leq$ `msw` $< 33$) <br /> $1 =$ Category 1 ($33  \leq$ `msw` $< 42$) <br /> $2 =$ Category 2 ($42 \leq$ `msw` $< 49$) <br /> $3 =$ Category 3 ($49 \leq$ `msw` $< 58$) <br /> $4 =$ Category 4 ($58 \leq$ `msw` $< 70$) <br /> $5 =$ Category 5 (`msw` $\ge 70$)| 5 | Optional$^{4}$  |

$^{1}$ Providing the `basin` allows filter storms on a particular basin 
using the `basin` argument of the `defDatabase` function. <br />

$^{2}$ Providing `rmw` allows better estimates of wind speed and direction
in the `temporatlBehaviour()` and `spatialBehaviour()` functions. If
not provided `rmw` is approximated using an empirical formula derived from
Willoughby et al. (2006) as follows, $R_m = 46.4e^{(-0.0155 \times v_m + 0.0169 \times
|\phi|)}$. <br />

$^{3}$ If the `pressure` and `poci` fields are not provided, both Holland (1980) and Boose et al. (2004) 
wind field models cannot be used in the `temporatlBehaviour()` and `spatialBehaviour()` functions. <br />

$^{4}$ The `sshs` field is optional, if not provided this field is automatically
filled using the `msw` field.

### Downloading a data source

If not available, a storm track data set has to be downloaded. In the following
example we illustrate how a storm track data set can be downloaded on the
IBTrACKS website. Different ".nc" (NetCDF) files containing storm track data
sets can be downloaded here:
<https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/netcdf/>.
Please, follow the IBTrACS citation recommendations when using these datasets
(including citing Knapp et al., 2010 and Knapp et al., 2018, and providing the
access date for the later). For example, track data for all the storms that
occurred over the last three years over the different basin world wide can be
download on the working directory as follows,

```{r, eval = F}
download.file(url = "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/netcdf/IBTrACS.last3years.v04r00.nc", destfile = "./IBTrACS_ALL_3yrs.nc")
```

### Creating a `StormsDataset`

Once a storm track data set is available a `StormsDataset` object can be created
using the `defDatabase` function as follows,

```{r, eval=F}
sds <- defDatabase(filename = "IBTrACS_ALL_3yrs.nc", basin=NULL, verbose = FALSE)
str(sds)
length(sds@database$names)
```

This `StormsDataset` object contains track data for 344 storms. When using an IBTrACKS storm track data set, default arguments can be used. By default, field provided by USA agencies (starting with "usa_") are used, but other fields can be selected. For example, for the South Pacific basin ("SP"), we can choose to use data provided by the Australian bureau of meteorology (starting with "bom_") as follows,

```{r, eval = F}
sds <-defDatabase(filename = "IBTrACS_ALL_3yrs.nc",
                  fields = c(names = "name",
                             seasons = "season",
                             isoTime = "iso_time",
                             basin = "basin",
                             lon = "bom_lon",
                             lat = "bom_lat",
                             msw = "bom_wind",
                             rmw = "bom_rmw",
                             pressure = "bom_pres",
                             poci = "bom_poci"),
                  basin = "SP",
                  verbose = FALSE)
length(sds@database$names)
```

This `StormsDataset` object contains track data for 38 storms in the South
Pacific basin.

### Unit conversion

In the IBTrACS data sets speeds are given in knots ($knt$), distances in
nautical miles ($nm$), and pressure in millibar ($mb$) but the models
implemented in the `temporalBehaviour()` and `spatialBehaviour()` functions to
compute wind speed, wind direction, and summary statistics require data
following the international system of units (SI) or SI-derived units. Therefore,
by default, when using IBTrACS data sets, the `defDatabase()` function converts
knots ($knt$) to meter per second ($m.s^{-1}$), nautical miles ($nm$) to
kilometres ($km$), and millibar ($mb$) to Pascal ($pa$).

When using another source of data than IBTrACS, the `unit_conversion` argument of the
`defDatabase()` function can be used to convert the data to the desire units.
The following conversions are available,

For `msw`: <br />
`knt_to_ms`: converts knot in meter per second (default setting) <br />
`kmh_to_ms`: converts kilometre per hour in meter per second <br />
`mph_to_ms`: converts miles per hour in meter per second <br />
`None`: no conversion <br />

For `rmw`: <br />
`nm_to_ms`: converts nautical miles in kilometre (default setting) <br />
`None`: no conversion <br />

For both `pressure` and `poci`: <br />
`mb_to_pa`: converts millibar in Pascal (default setting) <br />
`b_to_pa`: converts bar in Pascal <br />
`atm_to_pa`: converts atmosphere in Pascal <br />
`psi_to_pa`: converts psi in Pascal <br />
`None`: no conversion <br />

### Test data set

A `test_dataset` is provided with the `StormR` package. This test data set comprises the
track data of nine storms that occurred near Vanuatu and New Caledonia between 2015-2016
and 2020-2021, respectively. This `test_dataset` is used by default in
the `Storms()` function if no other `StormsDataset` has been created with the `defDatabase()` function.

## Reference

* Boose, E. R., Serrano, M. I., & Foster, D. R. (2004). Landscape and regional impacts of hurricanes in Puerto Rico. Ecological Monographs, 74(2), Article 2. https://doi.org/10.1890/02-4057
 
* Holland, G. J. (1980). An Analytic Model of the Wind and Pressure Profiles in Hurricanes. Monthly Weather Review, 108(8), 1212–1218. https://doi.org/10.1175/1520-0493(1980)108<1212:AAMOTW>2.0.CO;2

* Knapp, K. R., Kruk, M. C., Levinson, D. H., Diamond, H. J., & Neumann, C. J. (2010). The International Best Track Archive for Climate Stewardship (IBTrACS). Bulletin of the American Meteorological Society, 91(3), Article 3. https://doi.org/10.1175/2009bams2755.1

* Knapp, K. R., H. J. Diamond, J. P. Kossin, M. C. Kruk, & Schreck, C. J. (2018) International Best Track Archive for Climate Stewardship (IBTrACS) Project, Version 4. [indicate subset used]. NOAA National Centers for Environmental Information. https://doi:10.25921/82ty-9e16 [2023-03-15]
 
* Willoughby, H. E., Darling, R. W. R., & Rahn, M. E. (2006). Parametric Representation of the Primary Hurricane Vortex. Part II: A New Family of Sectionally Continuous Profiles. Monthly Weather Review, 134(4), 1102–1120. https://doi.org/10.1175/MWR3106.1