---
title: "Models"
output: html_document
date: "`r Sys.Date()`"
vignette: >
   %\VignetteIndexEntry{"Models"}
   %\VignetteEngine{knitr::rmarkdown}
   %\VignetteEncoding{UTF-8}
---



``` {r setup, include = FALSE}
library(StormR)
library(terra)
library(rworldxtra)
data("countriesHigh")
```

The main functions of the `StormR` package, `spatialBehaviour()` and
`temporalBehaviour()`, allow to compute characteristics of the storm surface
wind field, as re-constructed from storm track data and a parametric cyclone
model. Three parametric models are implemented in this package: Holland (1980),
Willoughby et al. (2006), and Boose et al. (2004). The use of one model or the
other is defined  using the `method` argument in  `spatialBehaviour()` and
`temporalBehaviour()` functions.

The original Holland (1980) and Willoughby et al. (2006) models provide a
symmetrical wind field around the cyclone centre. However cyclonic winds are not
symmetric, and an order zero asymmetry is caused by the storm translation
(forward motion). We therefore suggest using an asymmetric version of the
parametric wind fields that takes into account storm motion. In the `StormR`
package the methods developed by Miyazaki et al. (1962) and Chen (1994) that
allow to take this asymmetry into account can be used to adjust the outputs of
the symmetrical models accordingly. These can be activated by using the
`asymmetry` argument of the `spatialBehaviour()` and `temporalBehaviour()`
functions. The model of Boose et al. (2004) is already an asymmetrical version
of the Holland (1980) model. Contrary to the Holland (1980) and Willoughby et
al. (2006), this model considers different parameter settings over water or over
lands.

By default the `spatialBehaviour()` and `temporalBehaviour()` functions use the
Willoughby et al. (2006) model adjusted using the Chen (1994) method.

### Holland (1980) symmetric wind field

The Holland (1980) model, widely used in the literature, is based on the
gradient wind balance in mature tropical cyclones. The wind speed distribution
is computed from the circular air pressure field, which can be derived from the
central and environmental pressure and the radius of maximum winds.

$$
v_r = \sqrt{\frac{b}{\rho} \times \left(\frac{R_m}{r}\right)^b \times (p_{oci} - p_c) \times e^{-\left(\frac{R_m}{r}\right)^b} + \left(\frac{r \times f}{2}\right)^2} - \left(\frac{r \times f}{2}\right)
$$
where, <br />
$v_r$ is the radial wind speed (in $m.s^{-1}$) <br />
$r$ is the distance to the eye of the storm (in $km$) <br />
$R_m$ is the radius of maximum sustained wind speed (in $km$) <br />
$p_c$ is the pressure at the centre of the storm (in $Pa$) <br />
$p_{oci}$ is the pressure at outermost closed isobar of the storm (in $Pa$) <br />
$\rho$ is the air density set to $1.15 kg.m^{-3}$ <br />
$f = 2 \times 7.29 \times 10^{-5} \sin(\phi)$ is the Coriolis force (in $N.kg^{-1}$, with $\phi$ being the latitude) <br />
$b = \frac{\rho \times e \times v_m^2}{p_{oci} - p_c}$ is the shape parameter, with $e$ being the base of natural logarithms ~2.718282 and $v_m$ the maximum sustained wind speed (in $m.s^{-1}$) <br />
<br />

### Willoughby et al. (2006) symmetric wind field

The Willoughby et al. (2006) model is an empirical model fitted to aircraft
observations. The model considers two regions: inside the eye and at external
radii, for which the wind formulations use different exponents to better match
observations. In this model, the wind speed increases as a power function of the
radius inside the eye and decays exponentially outside the eye after a smooth
polynomial transition across the eyewall (see also Willoughby 1995, Willoughby
et al. 2004).

$$
\left\{
\begin{aligned}
v_r &= v_m \times \left(\frac{r}{R_m}\right)^{n} \quad if \quad r < R_m \\
v_r &= v_m \times \left((1-A) \times e^{-\frac{|r-R_m|}{X1}} + A \times e^{-\frac{|r-R_m|}{X2}}\right) \quad if \quad r \geq R_m \\
\end{aligned}
\right.
$$
where, <br />
$v_r$ is the radial wind speed (in $m.s^{-1}$) <br />
$r$ is the distance to the eye of the storm (in $km$) <br />
$v_m$ is the maximum sustained wind speed (in $m.s^{-1}$) <br />
$R_m$ is the radius of maximum sustained wind speed (in $km$) <br />
$X1 = 287.6 - 1.942 \times v_m + 7.799 \times \ln(R_m) + 1.819 \times |\phi|$<br />
$X2 = 25$ <br />
$n = 2.1340 + 0.0077 \times v_m - 0.4522 \times \ln(R_m) - 0.0038 \times |\phi|$<br />
$A = 0.5913 + 0.0029 \times v_m - 0.1361 \times \ln(R_m) - 0.0042 \times |\phi|$ and $A\ge0$<br />
$\phi$ is the latitude of the center of the storm 
<br />

### Adding asymmetry to Holland (1980) and Willoughby et al. (2006) wind fields

The asymmetry caused by the translation of the storm can be added as follows,

$\vec{V} = \vec{V_c} + C \times \vec{V_t}$

where, <br />
$\vec{V}$ is the combined, asymmetric wind field <br />
$\vec{V_c}$ is symmetric wind field <br />
$\vec{V_t}$ is the translation speed of the storm <br />
$C$ is function of $r$, the distance to the eye of the storm (in $km$) <br />

Two formulations of C proposed by Miyazaki et al. (1962) and Chen (1994) are implemented.

#### Miyazaki et al. 1962
$C = e^{(-\frac{r}{500} \times \pi)}$
<br />

#### Chen 1994
$C = \frac{3 \times R_m^{\frac{3}{2}} \times r^{\frac{3}{2}}}{R_m^3 + r^3 +R_m^{\frac{3}{2}} \times r^{\frac{3}{2}}}$

where, <br />
$R_m$ is the radius of maximum sustained wind speed (in $km$) <br />
<br />

### Boose et al. (2004) asymmetric model

The Boose et al. (2004) model, or “HURRECON” model, is a modification of the
Holland (1980) model (see also Boose et al., 2001). In addition to adding
asymmetry, this model treats of water and land differently, using different
surface friction coefficient for each.

#### Wind speed

Wind speed is computed as follows,

$$
v_r = F\left(v_m - S \times (1 - \sin(T)) \times \frac{v_h}{2} \right) \times \sqrt{\left(\frac{R_m}{r}\right)^b \times e^{1 - \left(\frac{R_m}{r}\right)^b}}
$$

$v_r$ is the radial wind speed (in $m.s^{-1}$) <br />
$v_h$ is the storm velocity (in $m.s^{-1}$) <br />
$r$ is the distance to the eye of the storm (in $km$) <br />
$v_m$ is the maximum sustained wind speed (in $m.s^{-1}$) <br />
$R_m$ is the radius of maximum sustained wind speed (in $km$) <br />
$p_c$ is the pressure at the centre of the storm ($pressure$ in $Pa$) <br />
$p_{oci}$ is the pressure at outermost closed isobar of the storm (in $Pa$) <br />
$\rho = 1.15$ is the air density (in $kg.m^{-3}$) <br />
$b = \frac{\rho \times e \times v_m^2}{p_{oci} - p_c}$ is the shape parameter <br />
$F$ is a scaling parameter for friction ($1.0$ in water, $0.8$ in land) <br /> 
$S$ is a scaling parameter for asymmetry ($1.0$) <br /> 
$T$ oriented angle (clockwise/counter clockwise in Northern/Southern Hemisphere) between forward trajectory of the storm and a radial line from the eye of the storm to point $r$ <br /> 
$S$ Asymmetry coefficient (usually set to 1) <br />

#### Wind direction

Wind direction is computed as follows,

$$
\left\{
\begin{aligned}
D = A_z - 90 - I \quad if \quad \phi > 0 \quad(Northern \quad Hemispher) \\
D = A_z - 90 - I \quad if \quad \phi \leq 0 \quad(Southern \quad Hemispher) \\
\end{aligned}
\right.
$$
where <br />
$D$ is the direction of the radial wind <br />
$A_z$ is the azimuth from point r to the eye of the storm <br />
$I$ is the cross isobar inflow angle ($20$ in water, $40$ in land) <br />
$\phi$ is the latitude <br />

### Wind fields comparison

Here, we compare wind fields generated by different models that can be used in
`StormR` for the same time and location (tropical cyclone Pam near Vanuatu)

```{r}
st <- Storms(loi = c(168.33,-17.73), names = "PAM",verbose=0)
PAM<-getObs(st,name="PAM")
par(mfrow=c(1,2))
pf <- spatialBehaviour(st,product="Profiles",method="Holland",asymmetry="None",verbose=0)
terra::plot(pf$PAM_Speed_41,main="Holland (1980)",cex.main=0.8,range=c(0,90))
terra::plot(countriesHigh,add=TRUE)
lines(PAM$lon,PAM$lat,lty=3)
pf <- spatialBehaviour(st,product="Profiles",method="Willoughby",asymmetry="None",verbose=0)
terra::plot(pf$PAM_Speed_41,main="Willoughby et al. (2006)",cex.main=0.8,range=c(0,90))
terra::plot(countriesHigh,add=TRUE)
lines(PAM$lon,PAM$lat,lty=3)
par(mfrow=c(1,2))
pf <- spatialBehaviour(st,product="Profiles",method="Holland",asymmetry="Miyazaki",verbose=0)
terra::plot(pf$PAM_Speed_41,main="Holland (1980) + Miyazaki et al. (1962)",cex.main=0.8,range=c(0,90))
terra::plot(countriesHigh,add=TRUE)
lines(PAM$lon,PAM$lat,lty=3)
pf <- spatialBehaviour(st,product="Profiles",method="Willoughby",asymmetry="Miyazaki",verbose=0)
terra::plot(pf$PAM_Speed_41,main="Willoughby et al. (2006) + Miyazaki et al. (1962)",cex.main=0.8,range=c(0,90))
terra::plot(countriesHigh,add=TRUE)
lines(PAM$lon,PAM$lat,lty=3)
par(mfrow=c(1,2))
pf <- spatialBehaviour(st,product="Profiles",method="Holland",asymmetry="Chen",verbose=0)
terra::plot(pf$PAM_Speed_41,main="Holland (1980) + Chen (1994)",cex.main=0.8,range=c(0,90))
terra::plot(countriesHigh,add=TRUE)
lines(PAM$lon,PAM$lat,lty=3)
pf <- spatialBehaviour(st,product="Profiles",method="Willoughby",asymmetry="Chen",verbose=0)
terra::plot(pf$PAM_Speed_41,main="Willoughby et al. (2006) + Chen (1994)",cex.main=0.8,range=c(0,90))
terra::plot(countriesHigh,add=TRUE)
lines(PAM$lon,PAM$lat,lty=3)
par(mfrow=c(1,2))
pf <- spatialBehaviour(st,product="Profiles",method="Boose",verbose=0)
terra::plot(pf$PAM_Speed_41,main="Boose et al. (2004)",cex.main=0.8,range=c(0,90))
terra::plot(countriesHigh,add=TRUE)
lines(PAM$lon,PAM$lat,lty=3)
```

## References
* Boose, E. R., Chamberlin, K. E., & Foster, D. R. (2001). Landscape and Regional Impacts of Hurricanes in New England. Ecological Monographs, 71(1), Article 1. https://doi.org/10.1890/0012-9615(2001)071[0027:LARIOH]2.0.CO;2

* Boose, E. R., Serrano, M. I., & Foster, D. R. (2004). Landscape and regional impacts of hurricanes in Puerto Rico. Ecological Monographs, 74(2), Article 2. https://doi.org/10.1890/02-4057

*Chen, K.-M. (1994). A computation method for typhoon wind field. Tropic Oceanology, 13(2), 41–48.
 
* Holland, G. J. (1980). An Analytic Model of the Wind and Pressure Profiles in Hurricanes. Monthly Weather Review, 108(8), 1212–1218. https://doi.org/10.1175/1520-0493(1980)108<1212:AAMOTW>2.0.CO;2

* Miyazaki, M., Ueno, T., & Unoki, S. (1962). The theoretical investigations of typhoon surges along the Japanese coast (II). Oceanographical Magazine, 13(2), 103–117.

* Willoughby, H. E. (1995). Normal-Mode Initialization of Barotropic Vortex Motion Models. Journal of the Atmospheric Sciences, 52(24), 4501–4514. https://doi.org/10.1175/1520-0469(1995)052<4501:NMIOBV>2.0.CO;2

* Willoughby, H. E., & Rahn, M. E. (2004). Parametric Representation of the Primary Hurricane Vortex. Part I: Observations and Evaluation of the Holland (1980) Model. Monthly Weather Review, 132(12), 3033–3048. https://doi.org/10.1175/MWR2831.1

* Willoughby, H. E., Darling, R. W. R., & Rahn, M. E. (2006). Parametric Representation of the Primary Hurricane Vortex. Part II: A New Family of Sectionally Continuous Profiles. Monthly Weather Review, 134(4), 1102–1120. https://doi.org/10.1175/MWR3106.1

* Yan, D., & Zhang, T. (2022). Research progress on tropical cyclone parametric wind field models and their application. Regional Studies in Marine Science, 51, 102207. https://doi.org/10.1016/j.rsma.2022.102207
