---
title: "spatialBehaviour"
output: html_document
date: "2023-03-03"
---

``` {r}
library(StormR)
```

The `spatialBehaviour` function allows computing wind speed and direction for each cell of a regular grid (i.e., a raster) for a given tropical cyclone or set of tropical cyclones. The `product="Profiles"` argument allows producing 2D wind fields at any time. The `spatialBehaviour` function also allows to compute three associated summary statistics: the maximum sustained wind speed (`product="MSW"`), the power dissipation index (`product="PDI"`) and the duration of exposure to winds reaching defined speed thresholds along the life span of the cyclones (`product="Exposure"`). Then the `plotBehaviour` and the `writeRast` functions can be used to visualise and export the output products.

### Computing and plotting spatialBehaviour products

In the following example we compute the behaviour of winds generated by the topical cyclone Pam (2015) nearby the Vanuatu using the Willoughby et al. (2006) model with the Chen (1994) asymmetry (default settings). First, track data are extracted as follows: 

```{r}
st <- Storms(loi = "Vanuatu", names = "PAM")
plotStorms(st)
```

#### Wind profiles

Using track data and the `spatialBehaviour` function with the `product="Profiles` argument we can generate 2D wind fields at any time as follows:

```{r}
pf <- spatialBehaviour(st,product="Profiles",verbose=0)
pf
```

The function returns a `SpatRaster` object with two rasters, one for the wind speed and one for the wind direction, for each observation or interpolated observation. Rasters' names follow the following terminology, the name of the storm in capital letters, "Speed" or "Direction", and the index of the observation, separated by underscores. Note that actual observations have entire indices (e.g., 41, 42, ...) while interpolated observation have decimals (e.g., 41.1, 41.2, ...). Indeed, most tropical cyclone track data bases, such as IBTrACKS (Knapp et al., 2010), provide observations every 3 or 6 hours. Therefore to be able to compute wind fields every 1 hour, the observations are interpolated.

Wind speed and direction profiles at the 41^th^ observation can be plotted as follows:

```{r}
plotBehaviour(st,pf$PAM_Speed_41)
plotBehaviour(st,pf$PAM_Direction_41)
```

#### Summary statisics

The `spatialBehaviour` function can compute the different `products` (i.e., `"Profiles"`, `"MSW"`, `"PDI"`, `"Exposure"`) either separately or together. Here, we compute all three summary statistics together as follows:

```{r}
ss <- spatialBehaviour(st,product=c("MSW","PDI","Exposure"),verbose=0)
ss
```

The `spatialBehaviour` function returns a `SpatRaster` object with eight rasters: one for the maximum sustained wind speed (`"MSW"`), one for the power dissipation index (`"PDI"`), and one for each of the six defaults wind thresholds values set for the duration of exposure (`"Exposure"`).

```{r}
names(ss)
```

By default, the function returns the duration of exposure (in hours) to wind speeds above the thresholds used by the Saffir-Simpson hurricane wind scale (i.e., 18, 33, 42, 49, 58, and 70 $m.s^{-1}$, Simpson, 1974). This can be change using the `wind_threshold` argument.

The maximum sustained wind speed can be plotted as follows:

```{r}
plotBehaviour(st,ss$PAM_MSW)
```

The power dissipation index can be plotted as follows:

```{r}
plotBehaviour(st,ss$PAM_PDI)
```

The duration of exposure to wind stronger than 58 $m.s^{-1}$ (i.e., Saffir-Simpson's categories 4 and 5) can be plotted as follows:

```{r}
plotBehaviour(st,ss$PAM_Exposure_58)
```

### Spatio-temporal resolution

As in the WorldClim database (<https://worldclim.org/data/worldclim21.html>, Fick and Hijmans, 2017) four spatial resolutions are available. By default the spatial resolution is set to 2.5 min (~4.5 km at the equator) but a finer spatial resolution of 30 s (~1 km at the equator) and coarser spatial resolutions of 5 min (~9 km at the equator) or 10 min (~18.6 km at the equator) can be set using the `space_res` argument. The temporal resolution is set to 1 hour by default but finer spatial resolution of 0.75, 0.50, or 0.25 hour can be set using the `temp_res` argument. Maximum sustained wind speed can be computed at a 10 min spatial resolution and a 30 min temporal resolution as follows: 

```{r}
ss <- spatialBehaviour(st,product=c("MSW"),verbose=0,space_res="10min",temp_res=0.5)
plotBehaviour(st,ss$PAM_MSW)
```

### Exporting spatialBehaviour products

The `spatialBehaviour` function returns rasters stored in a `SpatRaster` object than can be exported either in ".tiff" of ".nc" (NetCDF) formats using the `writeRast` function. Here, we export the maximum sustained wind speed in the working directory as follows:

``` r
writeRast(ss$PAM_MSW, format=".tiff")
```

## References

* Chen, K.-M. (1994). A computation method for typhoon wind field. Tropic Oceanology, 13(2), 41–48. Scopus.

* Fick, S. E., & Hijmans, R. J. (2017). WorldClim 2: New 1-km spatial resolution climate surfaces for global land areas. International Journal of Climatology, n/a-n/a. https://doi.org/10.1002/joc.5086

* Knapp, K. R., Kruk, M. C., Levinson, D. H., Diamond, H. J., & Neumann, C. J. (2010). The International Best Track Archive for Climate Stewardship (IBTrACS). Bulletin of the American Meteorological Society, 91(3), Article 3. https://doi.org/10.1175/2009bams2755.1

* Simpson, R. H. (1974). The hurricane disaster potential scale. Weatherwise, 27, 169–186.

* Willoughby, H. E., Darling, R. W. R., & Rahn, M. E. (2006). Parametric Representation of the Primary Hurricane Vortex. Part II: A New Family of Sectionally Continuous Profiles. Monthly Weather Review, 134(4), 1102–1120. https://doi.org/10.1175/MWR3106.1
 