---
title: "Storms"
output: html_document
date: "`r Sys.Date()`"
---


```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.align = 'center',
                      fig.pos = 't',
                      fig.width = 6,
                      fig.asp = 0.6,
                      cache = FALSE,
                      eval = TRUE)
options(knitr.table.format = "latex")
```


``` {r setup, include = FALSE}
library(StormR)
```

The `Storms()` function allows to extract tropical cyclone track data for a given tropical cyclone or set of tropical cyclones nearby a given location of interest (`loi`). The `loi` can be defined using a country name, a specific point (defined by its longitude and latitude coordinates), or any user imported or defined spatial polygon shapefiles. By default only observations located within 300 km around the `loi` are extracted but this can be changed using the `max_dist` argument. Users can also extract tropical cyclones using the `name` of the storm or the `season` during which it occurred. If the `name` and the `season` arguments are not filled then the `Storms()` function get all tropical cyclone since the first cyclonic season of the database. Once the data are extracted the `plotStorms()` function can be used to visualize the trajectories and points of observation of tropical cyclones on a map.

In the following example we use the `test_dataset` dataset provided with the package. 

### Getting and ploting tropical cyclone track data

#### Nearby a country

We extract data on the tropical cyclone PAM (2015) nearby Vanuatu as follows:  

``` {r chunk1}
st <- Storms(loi = "Vanuatu", names = "PAM",verbose=0)
```

The `Storms()` function returns a `StormsList` object in which the first slot `@data` contain a list of `Storm` objects. With the above specification the `StormsList` contain only one `Storm` object corresponding to PAM and the track data can be obtained using the `getObs()` function as follows:

``` {r chunk2}
head(getObs(st,name="PAM"))
```

The number of observation and the indices of the observations can be obtained using the `getNbObs()` and `getInObs()` as follows:

``` {r chunk3}
getNbObs(st,name="PAM")
getInObs(st,name="PAM")
```

The data can be visualised on a map as follows:

``` {r chunk4}
plotStorms(st, labels=TRUE)
```

#### Nearby a point

We extract all tropical cyclones nearby Nouméa (longitude = 166.45, latitude = -22.27) between 2015 and 2021 as follows:

``` {r chunk5}
pt <- c(166.45,-22.27)
st <- Storms(loi = pt, seasons = c(2015,2021), verbose=0)
```

The number, the names, and the season of occurrence of the storms in the returned `StormsList` object can be obtained using the `getNbStorms()`, `getNames()`, and `getSeasons()` functions as follows:

``` {r chunk6}
getNbStorms(st)
getNames(st)
getSeasons(st)
```

We choose to plot track data only for the topical cyclone Niran using the `names` argument of the `plotStorms()` function as follows:

``` {r chunk7}
plotStorms(st, names="NIRAN", labels=TRUE)
```

The track data for Niran can also be extracted and stored in a new object using the `getStorm()` function as follows:

``` {r chunk8}
NIRAN<-getStorm(st,name="NIRAN")
getNames(NIRAN)
```

#### Nearby a user defined spatial polygon shapefiles 

We extract all tropical cyclones occurring between 2015 and 2021 nearby the New Caledonia exclusive economic zone using the `eezNC` shapefile provided with the `StormR` package as follows:

``` {r chunk9}
sp <- eezNC
st <- Storms(loi = eezNC, season=c(2015,2021),verbose=0)
```

Information about the spatial extent of the track data exaction can be obtained using the `getLOI()`, `getBuffer()`, and `getBufferSize()` functions as follows:

``` {r chunk10}
LOI<-getLOI(st)
Buffer<-getBuffer(st)
BufferSize<-getBufferSize(st)
plot(Buffer,lty=3,main=paste(BufferSize,"km buffer arround New Caledonian EEZ",sep=" "))
plot(LOI,add=TRUE)
```

The maximum category of each cyclone on the Saffir-Simpson hurricane wind scale can be obtained using the `getSSHS()` function as follows:

``` {r chunk11}
getSSHS(st)
```

We can only plot category 4 and 5 tropical cyclones using the `category` argument of the `plotStorms()` function as follows:

``` {r chunk12}
plotStorms(st, category = c(4,5), labels=FALSE)
```
