---
title: "StormR"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{StormR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.align = 'center',
                      fig.pos = 't',
                      fig.width = 6, 
                      fig.asp = 0.6,
                      cache = FALSE,
                      eval = TRUE)
options(knitr.table.format = "latex")
```


```{r setup, include = FALSE}
library(StormR)
```


# Introduction

This vignette teaches you how to use StormR package with illustrations of few
basics examples. The main characteristics and features of the package's
functionalities are enlighten but we strongly advise the reader to check on the
documentation for a deeper insight of the whole package.

# Basic workflow 

From now on, we will demonstrate what must/may be done step by step throughout 
the different sub sections, in order to use this package in its best possible 
way.

## Choose a StormsDataset

The first thing to do after loading the package is to initialize a database for
Tropical Depressions/Storms/Cyclones (TD/TS/TC). This will let the package know
which database the user wants to extract TD/TS/TC from. We higlhy recommand
using IBTRaCS databases as they provide the most extensive and relevant
informations about TD/TS/TC overall.
Thus, this operation is carried out calling the initDatabase function whose
prototype with the default setting is showed right here:
```{r initdatabase, echo = FALSE, eval=FALSE}
initDatabase(
  filename = system.file("extdata", "IBTrACS.SP.v04r00.nc", package = "StormR"),
  fields = c(basin = "basin", names = "name", seasons = "season", isoTime = "iso_time",
    lon = "usa_lon", lat = "usa_lat", msw = "usa_wind", sshs = "usa_sshs", rmw =
    "usa_rmw", pressure = "usa_pres", poci = "usa_poci"),
  basin = NULL,
  verbose = TRUE
)
```

filename input must contain the path to the database the user is interested in.
Currently, this package can only read netcdf files (.nc). If one is interested
in using a database with another format (usually csv), note that several tools
exist to convert file in the netcdf format. Also, a dictionary input (field)
will inform which are the dimensions in the netcdf file that contain the desired
data. Note that some fields are mandatory (names, seasons, lon, lat, isoTime,
msw) while others are recommended (rmw, sshs) or highly recommended (pressure,
poci, basin). Last but not least, a basin input can be used as a filter so that
only TD/TS/TC that occured in the specified basin are extracted from the
database. We strongly recommend using this input as it may considerably speed up
the next functions.


This function returns in fine a StormsDataset object, especially designed to
gathers all these above informations about the database.

In what follows, the StormsDataset used relies on the IBTrACS.SP.v04r00.nc. This
dataset gathers all TD/TS/TC that occured in the South Pacific ocean since 1980.
This dataset named IBTRACS_SP is actually the default used in the package's
functions (See next section).


## Get data associated with storms

Once the StormsDataset is loaded, the next step is to select storms we are
interested in. This operation is done using getStorms function. It collects the
TD/TS/TC coming from a StormsDataset (via sds input) over a certain Location Of
Interest (loi input). This area is extended using the max_dist input afterwards.
Default value is set to 300km. Note that sds and loi inputs are mandatory to
perform this function (See getStorms Documentation). It is also possible to
filter TD/TS/TC by their cyclonic seasons and/or names (season and input names).
Finally, storms with maximum wind speed lower than 18 m/s (Tropical Depressions
in the Saffir Simpson Hurricane Scale SSHS) can be ignored using remove_TD
logical input. Default value is set to TRUE.


Here are some basic usage of the getStorms function.
In this case, we get the data associated with the tropical cyclone Harold that 
hit Vanuatu in 2020. Note here that the loi represents a whole country. (See
getStorms documentation to get the full list of country available).
```{r getStorms1}
harold <- getStorms(loi = "Vanuatu", names = "HAROLD")
```

In this second example, we collect data for all tropical storms and cyclones
over the Exclusive Economic Zone of New Caledonia (eezNC) between 2000 and 2022.
The loi here is a sf object, but it can also be a shapefile.
```{r getStorms2}
sts.nc <- getStorms(loi = eezNC, seasons = c(2000,2022))
```

In this last example, we retrieve all data associated with tropical storms and 
cyclones that occured since 1980 around the point coordinate 188.17:-13.92 
(longitude, latitude decimal degree) within a 300km buffer. These coordinates 
are actually located in the American Samoa.
```{r getStorms3}
pt <- c(188.17,-13.92)
sts.pt <- getStorms(loi = pt)
```


## Access data

The getStorms function returns data collected from TD/TS/TC in a Storms object
especially designed for this purpose (See Storms class). Then , one can be
interested in getting basics informations from a Storms object initialized with
getStorms. However the structure of this object is quite complex and it can
rapidly become overwhelming trying to reach data on your own. Here are some
getters that will help you saving time to access data.


From now on, we demonstrate how to use them using the sts.nc Storms object
initialized right above. First of all, if you are interested in getting all the
storm names, just run the following getter:
```{r getNames}
getNames(sts.nc)
```

Also, for each storms in your Storms object, the following getters will
respectively return the cyclonic season and the maximum category reached in the
SSHS:
```{r getSeasons}
#Get cyclonic seasons
getSeasons(sts.nc)
#Get maximum reached category in SSHS
getSSHS(sts.nc)
```

This getter simply returns the number of storms provided in your Storms Object:
```{r getNbStorms}
getNbStorms(sts.nc)
```


In addition, the next 3 getters are useful to retrieve spatial informations on
the LOI of your Storms object. The first command will return the LOI converted
in sf format:
```{r getLOI}
getLOI(sts.nc)
```

This second command simply returns the size (in km) of the buffer used to extent
the LOI:
```{r getBufferSize}
getBufferSize(sts.nc)
```

Finally this third command provides the LOI extended with the buffer:
```{r getBuffer}
getBuffer(sts.nc)
```

One can also be interested in getting the overall informations about a
particular storm. This operation is achieved using the following getter:
```{r getStorm}
niran <- getStorm(sts = sts.nc, name = "NIRAN")
```

Note: If serveral storms share the same name, the cyclonic season must be
specified to differentiate them. For example, 2 storms named Evan are provided
within the sts.pt Storms object initialized in the first section. This first
command will then throw an error, as we did not specify which one we are
interested in.
```{r getStorm1}
#getStorm(s = sts.pt, name = "EVAN")
```

We thus tackle this issue using the next 2 commands:
```{r getStorm2}
evan1997 <- getStorm(s = sts.pt, name = "EVAN", season = 1997)
evan2013 <- getStorm(s = sts.pt, name = "EVAN", season = 2013)
```

All these getters are designed to retrieve general informations on first levels
of Storms objects. However we can go further into the object getting data of a 
particular storm. In that way, the following getters work with both Storms and 
Storm signature (See Storm class).

These commands are simply getNames, getSeasons and getSSHS calls for signature 
Storm object:
```{r get signature Storm}
getNames(niran)
#Equivalent to getNames(s = sts.nc, name = "NIRAN")
getSeasons(niran)
#Equivalent to getSeasons(s = sts.nc, name = "NIRAN")
getSSHS(niran)
#Equivalent to getSSHS(s = sts.nc, name = "NIRAN")
```


To conclude, these 3 last getters provide information about observations for a
particular storm. They respectively return the number of observations, all of
the observations and the indices of observations within the extended LOI for a
particular storm, here tropical cyclone Niran (2021) over New Caledonia:
```{r getNbobs}
getNbObs(niran)
#Equivalent to getNbObs(s= sts.nc, name = "NIRAN")
getObs(niran)
#Equivalent to getObs(s = sts.nc, name = "NIRAN")
getInObs(niran)
#Equivalent to getInObs(s = sts.nc, name = "NIRAN")
```



## Plot data associated with storms

An interesting feature of this package is the plotStorms function which let the
user plot track(s) of storm(s) provided in a Storms object over and beyond the
LOI, using different settings (See plotStorms documentation to see all the
available input). Here are some basic usages of this function.

In this example, we plot tropical cyclone Harold track over the Vanuatu 
alongside with the labeled observations. Default settings are used to plot
labels which are every 24h and on the right side of observations.
```{r plotStorms1}
plotStorms(harold, labels = TRUE)
```


In this second example, we plot tropical cyclone Erica (2003) and Cook (2017), 
over the EEZ of New Caledonia alongside with the labeled observations (In this
case every ??H).
```{r plotStorms2}
plotStorms(sts.nc, names = c("ERICA", "COOK"), labels = TRUE, by = 12)
```

In this last example, we plot every tropical cyclone that reached category 5 
(SSHS) around the American Samoa, alongside with the labeled observations.
```{r plotStorms3}
plotStorms(sts.pt, category = 5, labels = TRUE)
```


## Computing rasterized products

The most important feature provided by this package may be by far the
stormBehaviour_sp function. Given a Storms object, it allows the user to compute
rasterized product(s) for each storm over the LOI. The available products are
the Maximum Sustained Wind (MSW) which is the default, the Power Index
Dissipation (PDI), the hour exposition for wind greater than a wind speed
threshold (Exposure) and finally 2D wind speed/direction structures for each
observations (Profiles). The output raster(s) are stacked in a SpatRaster object in WGS84
projection.

Depending on several inputs, the computations are not undertaken the same. The
user must specify a method used to regenerate the wind speed structures in order
to compute the desired product, but also the asymmetry used and other
parameters. To do so, the package provides two methods which are the wind models
derived from Holland (1980) REF? and Willoughby et al. (2006) REF? which is the
default. The first one relies on both basic cyclonic Physics and parameters
fitting according to cyclonic observations while the second one is based on fits
performed on cyclonic observations. As TD/TS/TC are usually not symmetrical, it
is possible to modify the structure of wind speed according to the formula
derived in Boose et al. (2001) REF ?

Outputs product may also differ modifying the following inputs. Spatial
resolution for grid rasters can be chosen among 4 options: 30sec, 2.5min which
is the default, 5min and 10min degree. This choice has been made to match
Wordlclim data so that it should be easily compared ... ??? . Moreover, the time
resolution (in hour) used to interpolate observations in the forthcoming
computations can also be chosen among 4 options: 1 (60min) which is the default,
0.75 (45min) , 0.5 (30min) and 0.25 (15min). Note that the finer the
spatial/time resolution, the slower the routine.



```{r stormBehaviour_sp 1}
prod.harold <- stormBehaviour_sp(harold, product = c("MSW", "PDI", "Exposure"))
```

```{r stormBehaviour_sp 2}
prof.harold <- stormBehaviour_sp(harold, product = "Profiles")
```

## Computing point wise products

If one is interested in getting the above products computed on a set of point
coordinates rather than computing them on raster grids which takes obsiously
more time, then the stormBehaviour_pt function should be used. It is actually
the stormBehaviour_sp counterpart where products are computed on specific
locations and not rasterized. Here are the main differences between these two
functions:
* The user must specify the set of point coordinate to use in a data.frame where 
  longitude and latitude must be respectively stored in a "lon" and "lat" 
  column.
* MSW product is replaced with the Time Series (TS) product which is the 
  default. It generates wind speed time series at the given location and over
  the whole storm lifecycle.
* Profiles product is no longer available as it does not make sense pointwise.


```{r stormBehaviour_pt 1}
luganville.pt <- data.frame(lon = 167.1667 , lat = -15.5333)

ts.luganville <- stormBehaviour_pt(harold, points = luganville.pt)
```


## Visualize products

```{r PlotBehaviour 1}
plotBehaviour(harold, prod.harold[["HAROLD_MSW"]])
```


```{r PlotBehaviour 2}
plotBehaviour(harold, prod.harold[["HAROLD_PDI"]])
```


```{r PlotBehaviour 3}
plotBehaviour(harold, prod.harold[["HAROLD_Exposure_58"]])
```

## Save product

```{r writeRast 1}
writeRast(prod.harold, path = paste0(tempdir(),"/"))
```

